Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler

### 2019-05-04 draft_paper/draft_paper.tex
stash

\maketitle

@@

The Burrows-Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. The forwards transform is straightforward to describe: for
a string of length \(n\), form the \(n \times n\) rotaion matrix of all
cyclic shifts of the original string, sort it lexicographically, and
take the last column of the matrix. In addition, we note the index
where the original string appears in the sorted rotation matrix. As
shown in \Cref{fig:bw_ex}, applying the algorithm to the string
\var{abracadabra!} produces the new string \var{ard!rcaaabb} and
the index \var{3}. The transformation tends to produce runs like
the three \var{a}s or the two \var{b}s in \var{ard!rcaaabb},
and it is this feature that makes the resulting string easier to
compress. Although it is not obvious, the transformed string, together
with the index, provide enough information to recover the the original
string. The BWT is discussed more in \cref{sec:bwt}.

\begin{figure}
  \centering
  \begin{tt}
  \begin{tabular}{rc}
    0  & abracadabra! \\
    1  & bracadabra!a \\
    2  & racadabra!ab \\
    3  & acadabra!abr \\
    4  & cadabra!abra \\
    5  & adabra!abrac \\
    6  & dabra!abraca \\
    7  & abra!abracad \\
    8  & bra!abracada \\
    9  & ra!abracadab \\
    10 & a!abracadabr \\
    11 & !abracadabra
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{rc}
    0  & !abracadabr\textbf{a} \\
    1  & a!abracadab\textbf{r} \\
    2  & abra!abraca\textbf{d} \\
    \textit{3} & \textit{abracadabra\textbf{!}} \\
    4  & acadabra!ab\textbf{r} \\
    5  & adabra!abra\textbf{c} \\
    6  & bra!abracad\textbf{a} \\
    7  & bracadabra!\textbf{a} \\
    8  & cadabra!abr\textbf{a} \\
    9  & dabra!abrac\textbf{a} \\
    10 & ra!abracada\textbf{b} \\
    11 & racadabra!a\textbf{b}
  \end{tabular}
  \end{tt}
  \caption{BWT acting on \var{abracadabra!}}
  \label{fig:bw_ex}
\end{figure}

@@

Unforunately, neither of these two definitions will be particularly
useful for proving the theorems we would like to prove with them. The
best definition for stability does not come by analogy with
\var{Permutation}. The intuition is that for any two lists that are
stable permutations of each other, if we filter each list by
equivalence with any element, the resulting lists should be identical.
This is because they must have the same elements, since the original
lists are permutations of one another, and those elements must be in
the same order, since the original lists are specifically stable
permutations of one another.
\lstinputlisting[
  widthgobble=2, firstline=21, lastline=22,
]{../theories/BWT/Sorting/Stable.v}

@@

\section{Insertion Sort, Radixsort, and \texttt{lexsort}}
\label{sec:sort}

>>>>

\maketitle
\vfill

@@

The Burrows-Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. It accomplishes this by creating runs of repeated letters
in the produced string. Of course, a simple way to create runs is to
sort the string, but that would be useless for compression because the
original string would be unrecoverable.

The forwards transform is straightforward to describe: for
a string of length \(n\), form the \(n \times n\) rotaion matrix of all
cyclic shifts of the original string, sort it lexicographically, and
take the last column of the matrix. In addition, we note the index
where the original string appears in the sorted rotation matrix. As
shown in \Cref{fig:bw_ex}, applying the algorithm to the string
\var{abracadabra!} produces the new string \var{ard!rcaaabb} and
the index \var{3}. The transformation tends to produce runs like
the three \var{a}s or the two \var{b}s in \var{ard!rcaaabb},
and it is this feature that makes the resulting string easier to
compress. Although it is not obvious, the transformed string, together
with the index, provide enough information to recover the the original
string. The BWT is discussed more in \cref{sec:bwt}.

\begin{figure}[!hb]
  \centering
  \begin{tt}
  \begin{tabular}{rc}
    0  & abracadabra! \\
    1  & bracadabra!a \\
    2  & racadabra!ab \\
    3  & acadabra!abr \\
    4  & cadabra!abra \\
    5  & adabra!abrac \\
    6  & dabra!abraca \\
    7  & abra!abracad \\
    8  & bra!abracada \\
    9  & ra!abracadab \\
    10 & a!abracadabr \\
    11 & !abracadabra
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{rc}
    0  & !abracadabr\textbf{a} \\
    1  & a!abracadab\textbf{r} \\
    2  & abra!abraca\textbf{d} \\
    \textit{3} & \textit{abracadabra\textbf{!}} \\
    4  & acadabra!ab\textbf{r} \\
    5  & adabra!abra\textbf{c} \\
    6  & bra!abracad\textbf{a} \\
    7  & bracadabra!\textbf{a} \\
    8  & cadabra!abr\textbf{a} \\
    9  & dabra!abrac\textbf{a} \\
    10 & ra!abracada\textbf{b} \\
    11 & racadabra!a\textbf{b}
  \end{tabular}
  \end{tt}
  \caption{BWT acting on \var{abracadabra!}}
  \label{fig:bw_ex}
\end{figure}

@@

Unforunately, neither of these two definitions will be particularly
useful for proving the theorems we would like to prove with them. The
best definition for stability does not come by analogy with
\var{Permutation}. The intuition is that for any two lists that are
stable permutations of each other, if we filter each list by
equivalence with any element, the resulting lists should be identical.
This is because they must have the same elements, since the original
lists are permutations of one another, and those elements must be in
the same order, since the original lists are specifically stable
permutations of one another.
\lstinputlisting[
  widthgobble=2, firstline=21, lastline=22,
]{../theories/BWT/Sorting/StablePerm.v}

@@

\section{Radixsort}
\label{sec:sort}

As
